A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-07-05 bioc-workflow.Rmd
Section on thresholding

```{r qcplot, message=FALSE, warning=FALSE}
library("pRoloc")
plot2D(hl, fcol = NULL)
```

@@

```{r qcbx, fig.width = 12}
par(mfrow = c(1, 2))
o <- order(hyperLOPIT$Fraction.No)
plotDist(hyperLOPIT[, o], pcol = "#00000010")
lines(colMeans(exprs(hyperLOPIT[, o])), col = "red", type = "b")
boxplot(exprs(hyperLOPIT[, o]))
```

@@

Visualisation and data exploration is an important aspect of data analyses allowing one to shed light on data structure and patterns of interest. Using the `r Biocpkg("pRolocGUI")` package we can interactively visualise, explore and interrogate quantitative spatial proteomics data. The `r Biocpkg("pRolocGUI")` package is currently under active development and it relies on the `shiny` framework for reactivity and interactivity. The package currently distributes 3 different GUI's ("main" (default), "compare" or "compare") which are wrapped and launched by the `pRolocVis` function.

The "main" application is designed for exploratory data analysis and is divied into 3 tabs: (1) PCA, (2) Profiles and (3) Table selection. The default view upon loading is the PCA tab, which features a clickable interface and zoomable PCA plot with an interactive data table for displaying the quantitation information. Particular proteins of interest can be highlighted using the text search box. There is also an alternate profiles tab for visualisation of the protein profiles, which can be used to examine the patterns of proteins of interest. The Table selection tab provides an interface to control data table column selection. In the below code chunk we lauch the main app.

@@

The "compare" application is useful for examining two replicate experiments, or two experiments from different conditions, treatments etc. Calling the argument `app = "compare"` as per the below code chunk launches the app wherein two PCA plots are loaded side-by-side. Common proteins between the two data sets can be searched for, identified and highlighted. As per the main application there is also a protein profiles tab to allow one to look at the patterns of protein profiles of interest in each dataset. One key difference between the main and compare app when launching the app is that the compare app requires a `MSnSetList` as input to the `pRolocVis` function. For example, in the code chunk below we first create a `MSnSetList` of replicates 1 and 2 of the hyperLOPIT data.

@@

The final application "classify", has been designed to view machine learning classification results according to user-specified thresholds for the assignment of proteins to its sub-cellular location.

@@

```{r classify, eval=TRUE}
svmRes <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "SVM.marker.set")
```

@@

```{r setColors, include=FALSE}
setStockcol(paste0(getStockcol(), 70))
```
```{r plotSVM, eval=TRUE}
## set point size of each protein to be inversely proportional to the SVM score
ptsze <- exp(fData(svmRes)$svm.scores) - 1

## plot new predictions
plot2D(svmRes, fcol = "svm", cex = ptsze)
addLegend(svmRes, fcol = "svm", where = "bottomleft", bty = "n", cex = .5)
```

@@

It is common when applying a supervised classification algorithm to set a specific score cutoff on which to define new assignments, below which classifications are set to unknown/unassigned. This is because in a supervised learning setup proteins can only be predicted to be localised to one of the sub-cellular niches that appear in the labelled training data. We can not guaranette (and do not expect) that the whole class diversity to be represented in the training data as (1) finding markers that represent the whole diversity of the cell is challenging and (2) many sub-cellular niches contain too few proteins to train on (note: we recommend a minimum of 13 markers per sub-cellular class for stratified 80/20 partitioning and cross-validation).

Deciding on a threshold is not trivial as classifier scores are heavily dependent upon the classifier used and different sub-cellular niches can exhibit different score distributions. We recommend one to set class-specific score distributions. In the code chunk below we display a boxplot of the score distributions per organelle.

```{r thresholds}

To help examine these distributions there is also a dedicated interactive application in the `r Biocpkg("pRolocGUI")` package. To access this application:

```{r classifyApp}

To launch the classify application:

- Essential for protein localisation as we do not want to force all proteins in to one category (allow for multiple lcoalisations or localisations outside the training data set e.g. new novel compartments)

- `getPredictions` function allows you to pass a score threshold. Also `classify` application available to interactively view output of thresholds

>>>>

```{r qcplot, message=FALSE, warning=FALSE}
library("pRoloc")
plot2D(hyperLOPIT, fcol = NULL)
```

@@

```{r qcbx, fig.width = 12}
par(mfrow = c(1, 2))
o <- order(hyperLOPIT$Fraction.No)
plotDist(hyperLOPIT[, o], pcol = "#00000010")
lines(colMeans(exprs(hyperLOPIT[, o])), col = "red", type = "b")
boxplot(exprs(hyperLOPIT[, o]), las = 2)
```

@@

Visualisation and data exploration is an important aspect of data
analyses allowing one to shed light on data structure and patterns of
interest. Using the `r Biocpkg("pRolocGUI")` package we can
interactively visualise, explore and interrogate quantitative spatial
proteomics data. The `r Biocpkg("pRolocGUI")` package is currently
under active development and it relies on the `shiny` framework for
reactivity and interactivity. The package currently distributes 3
different GUI's ("main" (default), "compare" or "compare") which are
wrapped and launched by the `pRolocVis` function. In the below code
chunk we lauch the main app (note, we do not need to specify the
argument, `app = "main"` as it is the default).

@@

As diplayed in the screenshot above, the "main" application is
designed for exploratory data analysis and is divied into 3 tabs: (1)
PCA, (2) Profiles and (3) Table selection. The default view upon
loading is the PCA tab, which features a clickable interface and
zoomable PCA plot with an interactive data table for displaying the
quantitation information. Particular proteins of interest can be
highlighted using the text search box. There is also an alternate
profiles tab for visualisation of the protein profiles, which can be
used to examine the patterns of proteins of interest. The Table
selection tab provides an interface to control data table column
selection.

The "compare" application is useful for examining two replicate
experiments, or two experiments from different conditions, treatments
etc. The compare application is called by default if the input object
to `pRolocVis` is a `MSnSetList` of 2 `MSnSets`, but it can also be
specified by calling the argument `app = "compare"`. For example, in
the code chunk below we first create a `MSnSetList` of replicates 1
and 2 of the hyperLOPIT data, this is then passed to `pRolocVis`.

@@

The comparison app loads the two PCA plots side-by-side. Only common
proteins between the two data sets are displayed. As per the main
application proteins can be searched, identified and highlighted on
both PCA plots and in the dedicated profiles tab. One key feature of
the compare application is the ability to re-map the second dataset
onto the PCA data space of the first (reference) data set (see
`?pRolocVis` and the argument `remap = TRUE`). Using the first dataset
as the reference set, PCA is carried out on the first dataset and the
standard deviations of the principal components (i.e. the square roots
of the eigenvalues of the covariance/correlation matrix) and the
matrix of variable loadings (i.e. a matrix whose columns contain the
eigenvectors) are stored and then used to calculate the principal
components of the second dataset. Both datasets are scaled and
centered in the usual way. The first dataset appears on the left, and
the second re-mapped data appears on the right. The order of the first
(the reference data for remapping) and second dataset can be changed
through regeneration/re-ordering of the `MSnSetList` object.

The final application "classify", has been designed to view machine
learning classification results according to user-specified thresholds
for the assignment of proteins to its sub-cellular location, this is
discussed later in the subsection [thresholding](#thresholding) in the
[supervised machine learning section](#supervised-machine-learning).

@@

```{r classify, eval=TRUE}
svmres <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "SVM.marker.set")
```

@@

```{r setColors, include=FALSE}
setStockcol(paste0(getStockcol(), 70))
```
```{r plotSVM, eval=TRUE}
## set point size of each protein to be inversely proportional to the
ptsze <- exp(fData(svmres)$svm.scores) - 1

## plot new predictions
plot2D(svmres, fcol = "svm", cex = ptsze)
addLegend(svmres, fcol = "svm", where = "bottomleft", bty = "n", cex = .5)
```

@@

It is common when applying a supervised classification algorithm to
set a specific score cutoff on which to define new assignments, below
which classifications are set to unknown/unassigned. This is important
as in a supervised learning setup proteins can only be predicted to be
localised to one of the sub-cellular niches that appear in the
labelled training data. We can not guaranette (and do not expect) that
the whole class diversity to be represented in the labelled training
data as (1) finding markers that represent the whole diversity of the
cell is challenging (especially obtaining dual- and multiply-localised
protein markers) and (2) many sub-cellular niches contain too few
proteins to train on (we recommend a minimum of 13 markers per
sub-cellular class for stratified 80/20 partitioning and 5-fold
cross-validation). <!--Conversely, with defining markers one must also
be careful to not label too many markers, as this can lead to
overfitting and mis-assignment.-->

Deciding on a threshold is not trivial as classifier scores are
heavily dependent upon the classifier used and different sub-cellular
niches can exhibit different score distributions, as highlighted in
the boxplot below. We recommend users to set class-specific thresholds.
In the code chunk below we display a boxplot of the score distributions 
per organelle.

```{r thresholds, fig.height=8, fig.width=8}
## First remove the markers
preds <- unknownMSnSet(svmres, fcol = "SVM.marker.set")

## Plot a boxplot of the scores of each organelle
par(oma= c(10.5, 0, 0, 0))
boxplot(svm.scores ~ svm, data = fData(preds), ylab = "SVM scores", las = 2)
```

There are many ways to set thresholds and the choice of method will depend
on the biological question and experimental design at hand. One viable
approach in the frame of the above experimetal design would be to manually
set a FDR, say 5%, per organelle. To do this the user would examine 
the top scoring predictions for each organelle, and then set a threshold at
the score at which they achieve 5% of false assignments per organelle. 
The definintion of a false assignment would depend on the information 
available, for example, validity or lack of validity for the localisation
from another experiment as reported in the literature or a reliable database. 
If such information is not available, one crude method is to set a threshold  
per organelle by extracting the median or 3rd quantile score per organelle. 
For example, in the code chunk below, we use the `orgQuants` function
to extract the median organelle scores and then pass these scores to the 
`getPrediction` function to extract the new localisations that meet this
scoring criteria. Any sub-cellular predictions that fall below the specified
thresholds are labelled as unknown.

```{r orgquants}
ts <- orgQuants(svmres, 
                fcol = "svm", 
                scol = "svm.scores",
                mcol = "SVM.marker.set", 
                t = .5)

preds <- getPredictions(svmres, 
                        fcol = "svm",
                        scol = "svm.scores", 
                        mcol = "SVM.marker.set",
                        t = ts)
```

The output of `getPredictons` is the original `MSnSet` dataset with a new feature variable appended to the feature data called `fcol.pred` (i.e. in our case `svm.pred`) containing the prediction results. The results can also be visualied using `plot2D` function.

```{r plotres}
plot2D(preds, fcol = "svm.pred")
```

There is also a dedicated interactive application to help users examine these distributions in the `r Biocpkg("pRolocGUI")` package. This app can be launched via the `pRolocVis` function and specifying the argument `app = "classify"` along with the relevent `fcol`, `scol` and `mcol` which refer to the columns in the feature data that contain the new assignments, assignment scores and markers respectively (see also `fvarLabels(svmres)`).

```{r classifyApp, eval=FALSE}
library("pRolocGUI")
pRolocVis(svmres, app = "classify", 
          fcol = "svm", 
	        scol = "svm.scores",
          mcol = "SVM.marker.set")
```

![The classify application](./Figures/classify.png)

The data is loaded and displayed on a PCA plot and a boxplot is used to
display the classifier scores by data class. On the left there is
a sidebar panel with sliders to control the thresholds upon which classifications 
are made. There are two types of cut-off that the user can choose from: (1)
"Quantile" and (2) "User-defined". By default, when the application is launched
quatile scoring is selected and set to 0.5, the median. The 
class-specific score thresholds that correspond to selecting the desired
quantile are shown on as red dots on the boxplot. The assignments on the PCA
plot are also updated according to the selected threshold. The quantile 
threshold can be set by moving the corresponding quantile slider. If one
wished to set their own cut-offs the "User-defined" radio button must be selected
and then the sliders for defining user-specified scores become active and the
scores and highlighted on the boxplot by blue dots. For more information we refer
users to the [`r Biocpkg("pRolocGUI")` tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/pRolocGUI/inst/doc/pRolocGUI.html).
